Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-10 demystify.tex
updates

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've realized that \textbf{international} students, especially those in smaller countries or less well-known universities, often lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this book was intended for \emph{international students}, but has been expanded to include information for \emph{US domestic students} (\autoref{sec:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this book know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

While this book can be useful for all students, it would most benefit \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research opportunities).

Students from top schools with strong research programs and experience might already know most of the information given here. They might also receive guidance from their professors or grad students who have gone through the process. Students from smaller schools or countries, however, might not have access to such resources and know where to start. 
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply and succeed in CS PhD programs in the US.

This book also aims to explain the reasons behind the admission process. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} that tell you \emph{``what''} to do, but few explain the \emph{``why''}, e.g., why LoRs matter so much, why you should not draft your own LoR, why you should contact professors, etc. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

@@

\begin{table}
    \caption{MS vs. PhD}\label{tab:phd-vs-ms}
    \centering
    \small
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{MS} & \textbf{PhD} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & No research experience & Research experience \\
    Coursework Req & Yes & Yes (research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

@@

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years as mentioned in \autoref{sec:time}) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation, which usually takes much longer time than coursework.

@@

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive in research, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS has quite a few TAs for MS students, and my courses in the past two years have MS TAs.

@@

Because of the differences between MS and PhD programs, you should consider different factors when selecting and ranking MS programs. For example, you might want to consider factors such as location, industry connections, and job placement instead of research areas and advisers (\autoref{sec:schoolsandprofs}).

@@

\item \textbf{Ranking:} Unlike PhD programs where the adviser and research are perhaps the most important factor, in MS programs the ranking of the school is more important.  Thus, you should consider the ranking of the school in CS, especially in the specific area you're interested in.  For example, if you're interested in software engineering, you might want to consider schools with strong SWE programs.

@@

\begin{table}
  \centering
  \small
  \caption{The top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\
    11 & University of Maryland - College Park &  36& Ohio State \\
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\
    13 & Purdue University &38& Yale University \\
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University\red{$^*$} &42 & Boston University  \\
    18 & Princeton University\red{$^*$}  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\

%\autoref{tab:ranking} lists .
\chapter{History and Acknowledgement}\label{sec:ack}
\paragraph{History} This handbook was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most participants are students in non-CS fields or not in the US, and their answers are not always accurate and can lead to more confusion. So I thought it would be useful to have a handbook that is specific to CS PhD programs in the US from an insider perspective.

I started writing this book in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The book was initially intended for international students but has been generalized to all students.
The handbook is open-source and available on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.
\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). Both UNL and Mason are large R1 research universities in the US and have many international students in their CS graduate programs. My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

I have been involved the PhD admission process at Mason and UNL for many years. This allows me to have a good understanding of the process, the challenges students face, and what faculty are looking for. Currently I serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with MS admission---which is quite different than PhD as discussed in \autoref{chap:ms}). I also have been in the panels of PhD fellowships including NSF GRFP and DoD NDSEG, which allows me to provide some insights on external fellowhip applications (\autoref{sec:fellowships}).

>>>>

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've observed that \textbf{international students}, especially those in smaller countries or less well-known universities, often lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this book was intended for international students, but has been expanded to include information for \textbf{US domestic students} (\autoref{sec:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this book know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

While this book can be useful for all students, it would most benefit \textbf{international students} from \textbf{smaller countries and less well-known universities}.  It is also useful for \textbf{domestic students} from US universities with \textbf{no PhD programs or limited research opportunities}.

Students from top schools with strong research programs and experience might already know most of the information given here. They likely have received guidance from their professors or grad students who have gone through the process. Students from smaller schools or countries, however, might not have access to such resources and know where to start. 
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply and succeed in CS PhD programs in the US.

This book also aims to explain the \textbf{reasons behind the admission process}.
There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} that tell you \emph{``what''} to do, but few explain the \emph{``why''}, e.g., why LoRs matter so much, why you should not draft your own LoR, why you should contact professors, etc. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

@@

\begin{table}
    \caption{MS vs. PhD}\label{tab:phd-vs-ms}
    \centering
    \small
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{MS} & \textbf{PhD} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & No research experience & Research experience \\
    Coursework Req & Yes & Yes (but research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

@@

  \item Course requirements: MS has a specific number of courses that typically can be done in 2 years. You graduate with an MS when you're done with the courses. PhD also has coursework requirements, which are typically taken in your first 2 years  (\autoref{sec:time}). However, after (and also during) coursework, you focus mainly on research.  You graduate with a PhD when you have done enough research and written a dissertation, which usually takes much longer time than coursework.

@@

\subsection{Exceptions} 
While MS students are typically not funded and do not focus on research, there are always exceptions.
Some MS programs are fully funded, e.g., those that are research-oriented and have a strong research component.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS has quite a few TAs for MS students, and my courses in the past two years have MS TAs.
Finally, if you have a strong background and can demonstrate that you can be productive in research, then a prof. might be willing to take you as their RA.  Some of these students convert to PhD students after having this research experience.

@@

Because of the differences between MS and PhD programs (\autoref{sec:phd-vs-ms}), you should consider different factors when selecting and ranking MS programs. For example, you might want to consider factors such as location, industry connections, and job placement instead of research areas and advisers (\autoref{sec:schoolsandprofs}).

@@

\item \textbf{Ranking:} Unlike PhD programs where the adviser and research are likely the most important factor, in MS programs the ranking of the school is typically more important.  Thus, you should consider the ranking of the school in CS, especially in the specific area you're interested in.  For example, if you're interested in software engineering, you might want to consider schools with strong SWE programs.

@@

\begin{table}
  \centering
  \small
  \caption{The top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences. (\href{https://csrankings.org}{CSRankings}, Jan. 2025). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Northwestern University\\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Pennsylvania State University  \\
    3 & Univ. of California-San Diego & 27 & Univ. of California - Riverside \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Santa Barbara \\
    5 & MIT                            & 30 &   Duke University \\
    6 & University of Michigan - Ann Arbor\red{$^*$}  & 30 & Rutgers University\red{$^*$} \\
    7 & University of Washington    & 32& University of Utah\\
    8 & Univ. of California - Berkeley    &33 & George Mason University\red{$^*$}\\
    9 & Cornell University   &34&   Texas A\&M University\red{$^*$}  \\
    10 & University of Maryland - College Park & 34 &  Univ. of California - Santa Cruz \\
    11 & Stanford University  &  36& North Carolina State University \\
    12 & Northeastern University\red{$^*$} &37& Ohio State \\
    13 & Purdue University &37& University of Virginia  \\
    14 & New York University  &37 & Yale University \\
    14 & University of Texas at Austin &40& Boston University  \\
    16 & Princeton University\red{$^*$}&41 & Univ. Of California - Davis \\
    16 & University of Pennsylvania  &42 & Brown University \\
    18 & Columbia University\red{$^*$}  & 42& Harvard \\
    19 & Univ. of California - Los Angeles  & 42& University of Illinois at Chicago  \\
    20 &  University of Wisconsin - Madison\red{$^*$}  &45& Arizona State University\red{$^*$} \\
    21 &  University of Southern California &45& University at Buffalo\red{$^*$} \\
    22 & University of Massachusetts-Amherst\red{$^*$} &45& University of North Carolina\red{$^*$} \\
    23 & University of Chicago &48& Oregon State University\red{$^*$}\\
    24 & Stony Brook University\red{$^*$} &48& Rice University\\
    25 &  Univ. of California - Irvine&48& University of Colorado-Boulder \\
    & &48& University of Minnesota\\ 
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\

%\autoref{tab:ranking} lists .
\chapter{History and Acknowledgement}\label{sec:ack}
\paragraph{History} This book was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most participants are students in non-CS fields or not in the US, and their answers are typically not accurate and often lead to more confusion. So I thought it would be useful to have a handbook that is specific to CS PhD programs in the US from an insider perspective.

I started writing this book in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The book was initially intended for international students but has been generalized to all students (domestic students also get confused by the process!).
The book is open-source and available on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.
\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). Both UNL and Mason are large R1 research universities in the US and have many international students in their CS graduate programs. My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

I have been involved the PhD admission process at Mason and UNL for many years. This allows me to have a good understanding of the process, the challenges students face, and what faculty are looking for. Currently I am the program director of the MS program in Software Engineering at Mason (thus also have experience with MS admission---which is quite different than PhD as discussed in \autoref{chap:ms}). I also have served in the panels of PhD fellowships including NSF GRFP and DoD NDSEG, which allows me to provide some insights on external fellowhip applications (\autoref{sec:fellowships}).
